---
title: "Single cell multi-omics"
author: "Burnet Bioinformatics group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
bibliography: references.bib
csl: harvard.csl
---

## Introduction

Joint RNA and ATAC analysis using 10x multiomic technology.
This workflow is adapted from the [Signac guide](https://stuartlab.org/signac/articles/pbmc_multiomic).

Source code: TBA.

There are datasets in two batches, April and August.

```{r,libs}

library("irlba")
library("Signac")
library("Seurat")
library("hdf5r")
library("EnsDb.Hsapiens.v86")
library("BSgenome.Hsapiens.UCSC.hg38")
library("glmGamPoi")
library("SeuratDisk") #devtools::install_github("mojaveazure/seurat-disk")
library("qlcMatrix") #devtools::install_github("cysouw/qlcMatrix")

```

## Load data

Load the RNA and ATAC data.

Annotate genes by organism.

```{r,load1}

counts <- Read10X_h5("FLW_april/outs/filtered_feature_bc_matrix.h5")
fragpath <- "FLW_april/outs/atac_fragments.tsv.gz"

```

Annotate genes by organism.

```{r,organno1}

#Mycoplasma pneumoniae chromosome
MPstart="gene-AXA72_00005"
MPend="gene-AXA72_03910"
MPgenes <- rownames(counts$`Gene Expression`)[which(rownames(counts$`Gene Expression`) == MPstart) : which(rownames(counts$`Gene Expression`) == MPend)]
MPstart="dnaA"
MPend="gene-EXC43_RS04465"
MPplas1genes <- rownames(counts$`Gene Expression`)[which(rownames(counts$`Gene Expression`) == MPstart) : which(rownames(counts$`Gene Expression`) == MPend)]
MPstart="dnaN"
MPend="gene-EXC43_RS04315"
MPplas2genes <- rownames(counts$`Gene Expression`)[which(rownames(counts$`Gene Expression`) == MPstart) : which(rownames(counts$`Gene Expression`) == MPend)]
MPgenes_full <- unique(c(MPgenes,MPplas1genes,MPplas2genes))
str(MPgenes_full)
table(colSums(as.data.frame(counts$`Gene Expression`[rownames(counts$`Gene Expression`) %in% MPgenes_full,])))

#lambda
Lstart="nu1"
Lend="orf206b"
Lgenes <- rownames(counts$`Gene Expression`)[which(rownames(counts$`Gene Expression`) == Lstart) : which(rownames(counts$`Gene Expression`) == Lend)]
str(Lgenes)
table(colSums(as.data.frame(counts$`Gene Expression`[rownames(counts$`Gene Expression`) %in% Lgenes,])))

#HHV5
H5start="RL10"
H5end="RL8A"
H5genes <- rownames(counts$`Gene Expression`)[which(rownames(counts$`Gene Expression`) == H5start) : which(rownames(counts$`Gene Expression`) == H5end)]
str(H5genes)
table(colSums(as.data.frame(counts$`Gene Expression`[rownames(counts$`Gene Expression`) %in% H5genes,])))

#Human gammaherpesvirus 4 (EBV)
Estart="BNRF1"
Eend="LMP-1"
Egenes <- rownames(counts$`Gene Expression`)[which(rownames(counts$`Gene Expression`) == Estart) : which(rownames(counts$`Gene Expression`) == Eend)]
str(Egenes)
table(colSums(as.data.frame(counts$`Gene Expression`[rownames(counts$`Gene Expression`) %in% Egenes,])))

#Human gene expression
HUstart="MIR1302-2HG"
HUend="LINC00266-4P"
HUgenes <- rownames(counts$`Gene Expression`)[which(rownames(counts$`Gene Expression`) == HUstart) : which(rownames(counts$`Gene Expression`) == HUend)]
str(HUgenes)

mycolsums <- colSums(as.data.frame(counts$`Gene Expression`[rownames(counts$`Gene Expression`) %in% HUgenes,]))
str(mycolsums)
summary(mycolsums)

```

## Get annotations

```{r,anno1}

annotation <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
seqlevels(annotation) <- paste0('chr', seqlevels(annotation))
annotation

```

## Create a Seurat object

It will contain RNA and ATAC data.

```{r,seurat1}

pbmc <- CreateSeuratObject( counts = counts$`Gene Expression`, assay = "RNA")
head(pbmc)

pbmc[["ATAC"]] <- CreateChromatinAssay(counts = counts$Peaks,sep = c(":", "-"),
  fragments = fragpath,annotation = annotation)

pbmc

```


## ATAC Quality control

Compute per-cell quality metrics using DNA accessibility data.

```{r,qc1}

DefaultAssay(pbmc) <- "ATAC"

pbmc <- NucleosomeSignal(pbmc)
pbmc <- TSSEnrichment(pbmc)

VlnPlot(
  object = pbmc,
  features = c("nCount_RNA", "nCount_ATAC", "TSS.enrichment", "nucleosome_signal"),
  ncol = 4,
  pt.size = 0
)

```

Remove cells that are outliers for these metrics, as well as cells with low or unusually
high counts for either the RNA or ATAC assay.

```{r,filt1}

pbmc <- subset(
  x = pbmc,
  subset = nCount_ATAC < 100000 &
    nCount_RNA < 25000 &
    nCount_ATAC > 1000 &
    nCount_RNA > 1000 &
    nucleosome_signal < 2 &
    TSS.enrichment > 1
)
pbmc

```

## Peak calling

Using MACS2.
We will remove peaks on nonstandard chromosomes and in genomic blacklist regions.
Then quantify counts in each peak and create a new assay using the MACS2 peak set
and add it to the Seurat object

```{r,peak1}

peaks <- CallPeaks(pbmc)

peaks <- keepStandardChromosomes(peaks, pruning.mode = "coarse")
peaks <- subsetByOverlaps(x = peaks, ranges = blacklist_hg38_unified, invert = TRUE)

macs2_counts <- FeatureMatrix(
  fragments = Fragments(pbmc),
  features = peaks,
  cells = colnames(pbmc)
)

pbmc[["peaks"]] <- CreateChromatinAssay(
  counts = macs2_counts,
  fragments = fragpath,
  annotation = annotation
)

```

##  Gene expression data processing

We can normalize the gene expression data using SCTransform, and reduce the dimensionality using PCA.

```{r,rnanorm1}

DefaultAssay(pbmc) <- "RNA"
pbmc <- SCTransform(pbmc)
pbmc <- RunPCA(pbmc)

```

## DNA accessibility data processing

Here we process the DNA accessibility assay the same way we would process a scATAC-seq dataset,
by performing latent semantic indexing (LSI).

```{r,atacnorm1}

DefaultAssay(pbmc) <- "peaks"
pbmc <- FindTopFeatures(pbmc, min.cutoff = 5)
pbmc <- RunTFIDF(pbmc)
pbmc <- RunSVD(pbmc)

```

## Annotating cell types

To annotate cell types in the dataset we can transfer cell labels from an existing PBMC reference dataset using
tools in the Seurat package.
See the Seurat reference mapping vignette for more information.

We’ll use an annotated PBMC reference dataset from Hao et al. (2020), available for
[download](https://atlas.fredhutch.org/data/nygc/multimodal/pbmc_multimodal.h5seurat).

Note that the SeuratDisk package is required to load the reference dataset.
Installation instructions for SeuratDisk can be found here.

Load reference data.

```{r,refdata}

options(timeout=10000)

if (! file.exists("pbmc_multimodal.h5seurat")) {
download.file("https://atlas.fredhutch.org/data/nygc/multimodal/pbmc_multimodal.h5seurat",
  destfile="pbmc_multimodal.h5seurat")
}

reference <- LoadH5Seurat("pbmc_multimodal.h5seurat", assays = list("SCT" = "counts"), reductions = 'spca')

reference <- UpdateSeuratObject(reference)

DefaultAssay(pbmc) <- "SCT"

# transfer cell type labels from reference to query
transfer_anchors <- FindTransferAnchors(
  reference = reference,
  query = pbmc,
  normalization.method = "SCT",
  reference.reduction = "spca",
  recompute.residuals = FALSE,
  dims = 1:50
)

predictions <- TransferData(
  anchorset = transfer_anchors,
  refdata = reference$celltype.l2,
  weight.reduction = pbmc[['pca']],
  dims = 1:50
)

pbmc <- AddMetaData(
  object = pbmc,
  metadata = predictions
)

# set the cell identities to the cell type predictions
Idents(pbmc) <- "predicted.id"

```

## Joint UMAP visualization

Using the weighted nearest neighbor methods in Seurat v4, we can compute a joint neighbor graph that
represent both the gene expression and DNA accessibility measurements.

```{r,jointumap}

# build a joint neighbor graph using both assays
pbmc <- FindMultiModalNeighbors(
  object = pbmc,
  reduction.list = list("pca", "lsi"),
  dims.list = list(1:50, 2:40),
  modality.weight.name = "RNA.weight",
  verbose = TRUE
)

# build a joint UMAP visualization
pbmc <- RunUMAP(
  object = pbmc,
  nn.name = "weighted.nn",
  assay = "RNA",
  verbose = TRUE
)

DimPlot(pbmc, label = TRUE, repel = TRUE, reduction = "umap") + NoLegend()

```

## Linking peaks to genes

For each gene, we can find the set of peaks that may regulate the gene by by computing the correlation between gene
expression and accessibility at nearby peaks, and correcting for bias due to GC content, overall accessibility,
and peak size.
See the [Signac paper](https://www.biorxiv.org/content/10.1101/2020.11.09.373613v1) for a full description of the
method we use to link peaks to genes.

Running this step on the whole genome can be time consuming, so here we demonstrate peak-gene links for a subset of
genes as an example.
The same function can be used to find links for all genes by omitting the genes.use parameter.


```{r,peaklink}

DefaultAssay(pbmc) <- "peaks"

# first compute the GC content for each peak
pbmc <- RegionStats(pbmc, genome = BSgenome.Hsapiens.UCSC.hg38)

# link peaks to genes
pbmc <- LinkPeaks(
  object = pbmc,
  peak.assay = "peaks",
  expression.assay = "SCT"
)

```

We can visualize these links using the `CoveragePlot()` function, or alternatively we could use the `CoverageBrowser()`
function in an interactive analysis.

```{r,peak2geneviz}

idents.plot <- c("B naive", "B intermediate", "B memory",
                 "CD14 Mono", "CD16 Mono", "CD8 TEM", "CD8 Naive")

p1 <- CoveragePlot(
  object = pbmc,
  region = "MS4A1",
  features = "MS4A1",
  expression.assay = "SCT",
  idents = idents.plot,
  extend.upstream = 500,
  extend.downstream = 10000
)

p2 <- CoveragePlot(
  object = pbmc,
  region = "LYZ",
  features = "LYZ",
  expression.assay = "SCT",
  idents = idents.plot,
  extend.upstream = 8000,
  extend.downstream = 5000
)

patchwork::wrap_plots(p1, p2, ncol = 1)

```

## Gene activity - looking for pathogen gene expression

```{r,activity1}

gene.activities <- GeneActivity(pbmc)

# add the gene activity matrix to the Seurat object as a new assay and normalize it
pbmc[['RNA']] <- CreateAssayObject(counts = gene.activities)
pbmc <- NormalizeData(
  object = pbmc,
  assay = 'RNA',
  normalization.method = 'LogNormalize',
  scale.factor = median(pbmc$nCount_RNA)
)


```

Now we can visualize the activities of canonical marker genes.

```{r,markerviz1}

DefaultAssay(pbmc) <- 'RNA'

FeaturePlot(
  object = pbmc,
  features = c('MS4A1', 'CD3D', 'LEF1', 'NKG7', 'TREM1', 'LYZ'),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3
)

```

Now we can visualize the activities of pathogen genes.

CP014267.1 = Mycoplasma pneumoniae chromosome

NZ_LR214945.1 = Mycoplasmoides pneumoniae strain NCTC10119 plasmid 1

NZ_LR214946.1 = Mycoplasmoides pneumoniae strain NCTC10119 plasmid 2

NC_001416.1 = lambda phage

NC_006273.2 = Human herpesvirus 5

NC_007605.1 = Human gammaherpesvirus 4

```{r,markerviz2}

FeaturePlot(
  object = pbmc,
  features = c('MS4A1', 'CD3D', 'LEF1', 'NKG7', 'TREM1', 'LYZ'),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3
)

```

## Save data

```{r,save}

save.image("scmulti.Rdata")

```

## Session information

Show versions of packages and key dependancies.

```{r,sessioninfo}

sessionInfo()

```
